Reconstructing the neutron-star equation of state with gravitational-wave detectors from a realistic population of inspiralling binary neutron stars

### 2014-08-11 paper.tex
Changed figures according to Les' suggestions.

\begin{abstract}
Gravitational-wave observations of inspiralling binary neutron star systems can be used to measure the neutron-star equation of state (EOS) through the tidally induced shift in the waveform phase that depends on the tidal deformability parameter $\lambda$. Previous work has shown that $\lambda$, a function of the neutron-star EOS and mass, is marginally measurable by Advanced LIGO for a single event when including tidal information up to the merger frequency. In this work, we describe a method for stacking measurements of $\lambda$ from multiple inspiral events to measure the EOS. We use Markov chain Monte Carlo simulations to estimate the parameters of a 4-parameter piecewise polytrope EOS that matches theoretical EOS models to a few percent. We find that for ``realistic'' event rates ($\sim 40$ events per year with signal-to-noise ratio $> 8$ in a single Advanced LIGO detector), when a year of data is combined with the constraints from causality and recent high mass neutron-star measurements, the EOS above nuclear density can be measured to better than a factor of two in most cases. We also find that for masses $>1M_\odot$, the neutron-star radius can be measured to $\pm 1$~km and the tidal deformability can be measured to $\pm 1 \times 10^{36}$~g~cm$^2$~s$^2$. The overwhelming majority of this information comes from the loudest $\sim 5$ events. Current uncertainties in the post-Newtonian waveform model, however, lead to systematic errors in the EOS measurement that are as large as the statistical errors, and more accurate waveform models are needed to remove this error.
\end{abstract}

@@

@@

[[You don't define SNR or network SNR anywhere.]]

@@

@@

\begin{figure}[!htb]
\begin{center}
\includegraphics[width=3.2in]{Radiuslambda.pdf}
\caption{Radius and tidal deformability of tabulated EOS models (solid) and the least-squares piecewise-polytrope fits to those tabulated models (dashed). The 20 vertical lines represent the most likely NS masses of the 10 known BNS systems~\cite{Lattimer2012}. Some of these masses, however, have significant uncertainties. The overlapping vertical bands represent the $1\sigma$ uncertainty in the masses of the pulsars J1614-2230~\cite{DemorestPennucciRansom2010} and J0348+0432~\cite{AntoniadisFreireWex2013}, both in neutron star-white dwarf binaries.}
\label{fig:rlofm}
\end{center}
\end{figure}

@@

Given the GW model and EOS fit described in Sections~\ref{sec:waveform} and~\ref{sec:eos}, we will describe here a method for estimating its parameters from a set of GW observations. We use Bayes' theorem to write the posterior density function (PDF) for the universal EOS parameters $\vec E = (\log(p_1), \Gamma_1, \Gamma_2, \Gamma_3)$ common to all NSs and the waveform parameters $\vec\theta = (D_L, \alpha, \delta, \psi, \iota, t_c, \phi_c, m_1, m_2, \tilde\Lambda)$ \red{[[What happened to $\delta\tilde\Lambda$? Although $\delta\tilde\Lambda$ in principle provides additional EOS information, it's effect on the waveform is small compared to $\tilde\Lambda$, so we treat it as a nuisance parameter and group it with the extrinsic parameters when we marginalize over them.]]} that are unique to each of the $n$ BNS systems as
\begin{equation}
\begin{split}
p(\vec E,\vec\theta_1,\dots,\vec\theta_n | D,\mathcal{H},\mathcal{I})
&= p(\vec E,\vec\theta_1,\dots,\vec\theta_n | \mathcal{H},\mathcal{I}) \\
& \times\frac{ p(D | \vec E,\vec\theta_1,\dots,\vec\theta_n,\mathcal{H},\mathcal{I}) }{ p(D | \mathcal{H},\mathcal{I}) }.
\end{split}
\end{equation}
The data $D = \vec d_1,\dots, \vec d_n$ is composed of the data streams $\vec d_i(t)$ from the GW detector network for each of the $n$ inspiral events. $\mathcal{H}$ represents the EOS and waveform models, and $\mathcal{I}$ represents the background information for the parameters. The quantity $p(\vec E,\vec\theta_1,\dots,\vec\theta_n | \mathcal{H},\mathcal{I})$ is the prior probability density for the EOS and waveform parameters, $p(D | \vec E,\vec\theta_1,\dots,\vec\theta_n,\mathcal{H},\mathcal{I})$ is the likelihood, and the normalization constant $p(D | \mathcal{H},\mathcal{I})$ is the evidence. The marginalized PDF for the EOS parameters is then
\begin{equation}
\label{eq:margEOS}
p(\vec E | D,\mathcal{H},\mathcal{I}) = \int d\vec\theta_1 \dots d\vec\theta_n p(\vec E,\vec\theta_1,\dots,\vec\theta_n | D,\mathcal{H},\mathcal{I}).
\end{equation}
This $10n$-dimensional integral is not easily computed, so we will now rewrite it. First we decompose the waveform parameters into extrinsic parameters $\vec\theta_{\rm ex} = (D_L, \alpha, \delta, \psi, \iota, t_c, \phi_c)$ that depend on the observer and intrinsic parameters $\vec\theta_{\rm in} = (m_1, m_2, \tilde\Lambda)$ that are relevant for measuring the EOS such that $\vec\theta = (\vec\theta_{\rm ex}, \vec\theta_{\rm in})$. 
%Although at this stage we treat the EOS parameters separately from the waveform parameters, we will later use the fact that the tidal parameter $\tilde\Lambda_i$ for system $i$ is a function $\tilde\Lambda_i = \tilde\Lambda(m_{1i}, m_{2i}, \vec E)$ of the masses of system $i$ and the EOS parameters to determine the EOS parameters.

@@

The marginalized PDF (Eq.~\eqref{eq:margEOS}) is now
\begin{equation}
\begin{split}
p(\vec E | D,\mathcal{H},\mathcal{I}) = &\frac{1}{p(D | \mathcal{H},\mathcal{I})} 
\int d\vec\theta_{\rm in, 1} \dots d\vec\theta_{\rm in, n} \left\{ p(\vec E | \mathcal{H},\mathcal{I}) \right. \\
& \times \prod_{i=1}^n \left[p(m_{1i}, m_{2i} | \vec E, \mathcal{H},\mathcal{I}) p(\tilde\Lambda_i | m_{1i}, m_{2i}, \vec E, \mathcal{H},\mathcal{I}) \right.\\
& \left.\left. \times L(\vec d_i; \vec\theta_{\rm in, i}, \mathcal{H}, \mathcal{I}) \right]\right\},
\end{split}
\end{equation}
where we have defined the quasilikelihood for the intrinsic parameters as
\begin{equation}
\label{eq:quasilike}
L(\vec d_i; \vec\theta_{\rm in, i}, \mathcal{H}, \mathcal{I}) = \int d\vec\theta_{\rm ex, i} p(\vec\theta_{\rm ex, i} | \mathcal{H}, \mathcal{I}) p(\vec d_i | \vec\theta_i,\mathcal{H},\mathcal{I}).
\end{equation}
Because $\tilde\Lambda_i$ is a deterministic function of $m_{1i}$, $m_{2i}$ and the EOS parameters,
\begin{equation}
p(\tilde\Lambda_i | m_{1i}, m_{2i}, \vec E, \mathcal{H},\mathcal{I}) = \delta(\tilde\Lambda_i - \tilde\Lambda(m_{1i}, m_{2i}, \vec E)).
\end{equation}
The marginalized PDF then finally becomes
\begin{equation}
\label{eq:margEOS2}
\begin{split}
p(\vec E | D,\mathcal{H},\mathcal{I}) =& \frac{1}{p(D | \mathcal{H},\mathcal{I})} \int dm_{11} dm_{21} \dots dm_{1n} dm_{2n} \\
&\times p(\vec E | \mathcal{H},\mathcal{I}) \prod_{i=1}^n \left[ p(m_{1i}, m_{2i} | \vec E, \mathcal{H},\mathcal{I}) \right.\\
& \left. \times L(\vec d_i; \vec\theta_{\rm in, i}, \mathcal{H}, \mathcal{I})|_{\tilde\Lambda_i = \tilde\Lambda(m_{1i}, m_{2i}, \vec E)} \right].
\end{split}
\end{equation}

@@

\subsection{Likelihood}

The final ingredient we need to evaluate the marginalized PDF is an expression for the likelihood $p(\vec d_i | \vec\theta_i,\mathcal{H},\mathcal{I})$ for each GW event. In this paper we assume that each detector in the network has stationary, Gaussian noise and that the noise between detectors is uncorrelated. This means that the power spectral density (PSD) $S_n(f)^a$ of the noise $n^a(t)$ in detector $a$ is 
\begin{equation}
\langle \tilde n^a(f) \tilde n^{b*}(f') \rangle = \frac{1}{2} \delta^{ab} \delta (f-f') S_n(f)^a,
\end{equation}
where $\tilde n^a(f)$ is the Fourier transform of the noise of detector $a$, and $\langle\cdot\rangle$ represents an ensemble average. For a GW event with true parameters $\hat\theta$, resulting in the gravitational wave signal $h^a(t; \hat\theta)$, the data stream of detector $a$ will be
\begin{equation}
d^a(t) = n^a(t) + h^a(t; \hat\theta).
\end{equation}

For stationary, Gaussian noise, it is well known that the probability of obtaining the noise time series $n(t)$ is
\begin{equation}
p_n[n(t)] \propto e^{-(n, n)/2},
\end{equation}
where $(a, b)$ is the usual inner product between two time series $a(t)$ and $b(t)$ weighted by the PSD
\begin{equation}
(a, b)=4{\rm Re} \int_0^\infty \frac{\tilde a(f) \tilde b^*(f)}{S_n(f)}\,df.
\end{equation}
If we have a GW model $m(t;\vec\theta)$ that approximates the true signal $h(t;\vec\theta)$, then the likelihood of obtaining the data $d^a(t)$ given that GW model is
\begin{equation}
p(d^a | \vec\theta, \mathcal{H}, \mathcal{I}) = p_n[d^a(t)-m(t; \vec\theta)] \propto e^{-(d^a - m(\vec\theta), d^a - m(\vec\theta))/2}.
\end{equation}
If the model $m$ differs from the true signal $h$, a systematic error will be introduced in the recovered waveform parameters. For a network of independent detectors described by the set of time series $\vec d(t)$, the likelihood is then
\begin{equation}
p(\vec d | \vec\theta, \mathcal{H}, \mathcal{I}) = \prod_a p(d^a | \vec\theta, \mathcal{H}, \mathcal{I}).
\end{equation}

We evaluate Eq.~\eqref{eq:margEOS2} for the marginalized PDF in a two-step procedure similar to that described by Steiner {\it et al.}~\cite{SteinerLattimerBrown2010}, where we first evaluate the quasilikelihood (Eq.~\eqref{eq:quasilike}) for each of the $n$ BNS systems, then evaluate Eq.~\eqref{eq:margEOS2}. In the first step where we evaluate the quasilikelihood $L(\vec d_i; \vec\theta_{\rm in, i}, \mathcal{H}, \mathcal{I})$, we use the MCMC sampler \texttt{lalinference\_mcmc} included in the LSC Algorithm Library~\cite{lal}, which is specifically designed to efficiently sample the BNS parameter space\red{[[cite overview papers]]}. The priors on the extrinsic parameter are~\red{[[what are they?]]}. The priors for the intrinsic parameters $(m_{1i}, m_{2i}, \tilde\Lambda_i)$ are not contained in the quasilikelihood (Eq.~\eqref{eq:quasilike}), so we use a flat prior for the intrinsic parameters. For each binary event $i$ we perform 10 independent runs and test for convergence using the Gelman-Rubin diagnostic~\cite{GelmanRubin1992}, then join the samples. Marginalization over the extrinsic parameters is performed by generating a distribution from the samples of the three intrinsic parameters with a Gaussian kernel density estimator. In the second step, we sample the $4+2n$ parameter integrand of Eq.~\eqref{eq:margEOS2} using the affine-invarient ensemble sampler \texttt{emcee}~\cite{ForemanMackeyHogg2013}. We use the EOS priors stated in Section~\ref{sec:eos}, and uniform priors for the masses with $0.5M_\odot \le m_2 \le m_1 \le 3.0M_\odot$. We again perform 10 runs and test for convergence with the Gelman-Rubin diagnostic before joining the samples. The marginalized distribution $p(\vec E|D, \mathcal{H}, \mathcal{I})$ is then given by a histogram of the samples for $\vec E$.

The evaluation of the quasilikelihood (Eq.~\eqref{eq:quasilike}) for all $n$ events in a population is very computationally expensive. When we examine many different populations in Section V we will sometimes use the Fisher matrix approximation instead. In the large SNR limit, the difference $\Delta\vec\theta = \vec\theta - \hat\theta$ between the estimated parameters $\vec\theta$ and the true parameters $\hat\theta$ of the binary system obeys a Gaussian distribution~\cite{FinnChernoff1993}. Specifically, for $N$ parameters, the likelihood is
\begin{equation}
\label{eq:gaussianlike}
p(d^a | \vec\theta,\mathcal{H},\mathcal{I}) = \frac{1}{\sqrt{(2\pi)^N \mathrm{det}(\Sigma_{ij})}} e^{-\frac{1}{2} \Sigma^{-1}_{ij} \Delta\theta_i \Delta\theta_j},
\end{equation}
where $\Sigma_{ij}$ is the covariance matrix, and it is given in terms of the Fisher matrix
\begin{equation}
\Gamma_{ij} = (\partial_i h(\hat\theta), \partial_j h(\hat\theta))
\end{equation}
by the relation $\Sigma_{ij}  = \Gamma^{-1}_{ij}$. In the large SNR limit, $\hat\theta$ will be approximately given by the maximum likelihood. When we use the Fisher matrix approximation we will use the mass variables $\ln(\mathcal{M})$ and $\ln(\eta)$ instead of $m_1$ and $m_2$ and flat priors for $\ln(\mathcal{M})$ and $\ln(\eta)$. The quasilikelihood (Eq.~\eqref{eq:quasilike}) marginalized over the extrinsic parameters is simply given by the submatrix of $\Sigma_{ij}$ containing the intrinsic parameters $(\ln(\mathcal{M}), \ln(\eta), \tilde\Lambda)$.

@@

In order to determine the characteristic ability of GW detectors to measure EOS parameters, we will want to average our results for Eq.~\eqref{eq:margEOS2} over many realizations of the detector noise. In Ref.~\cite{NissankeHolzHughes2010} it was shown that averaging over noise realizations is equivalent to setting the noise realization $n^a(t)$ used to generate the data $d^a$ equal to zero, but still using the characteristic noise PSD in the expression for the likelihood. Explicitly, one can define the averaged likelihood as the geometric mean of a set of $M$ identical detectors measuring the same gravitational wave event
\begin{equation}
p_{\rm ave}(d^a | \vec\theta, \mathcal{H}, \mathcal{I}) = \left[\prod_{k=1}^M p(d_k^a | \vec\theta, \mathcal{H}, \mathcal{I})\right]^{1/M}.
\end{equation}
Taking $M$ to be large, it was shown that
\begin{equation}
\label{eq:averagedlike}
p_{\rm ave}(d^a | \vec\theta, \mathcal{H}, \mathcal{I}) \propto e^{-\frac{1}{2}(m^a(\theta) - h^a(\hat\theta) | m^a(\theta) - h^a(\hat\theta))}.
\end{equation}
This averaged likelihood will allow us to examine characteristic measurement uncertainties in the EOS parameters independent of the particular noise realization. It will also allow us in Section~VI to separate the effects of systematic errors due to uncertainties in the waveform model from effects due to individual noise realizations.

@@

\begin{figure*}[!htb]
\begin{center}
\includegraphics[width=3.2in]{LALMCMCmpa1FitTaylorF2ZeroNoiseBNS20p.pdf}
\includegraphics[width=3.2in]{LALMCMCmpa1FitTaylorF2ZeroNoiseBNS20Radiuslambda.pdf}
\caption{Uncertainty in the recovered EOS, radius, and tidal deformability for the three-detector aLIGO--aVirgo network. Results are shown for the loudest 20 events with Network SNR from 64.1 down to 13.7. The red, green, and blue regions represent the 68\% ($1\sigma$), 95\% ($2\sigma$), and 99.7\% ($3\sigma$) confidence regions respectively. In the bottom left panel, $p_{\rm true}$ is the pressure of the true ``injected'' EOS, which in this case is the fit to the MPA1 EOS. The tabulated MPA1 EOS is also shown. The range of masses from $1.2M_\odot$ to $1.6M_\odot$ used in the simulated BNS population is represented in the right panels by the black vertical dotted lines. The red vertical dotted line at $1.93M_\odot$ is the minimum mass, set by the prior, where some accepted EOS parameters do not produce a stable NS.}
\label{fig:mpa1fit}
\end{center}
\end{figure*}

@@

[[Thank Andrew Steiner for discussion at INT.]]

>>>>

\begin{abstract}
Gravitational-wave observations of inspiralling binary neutron star systems can be used to measure the neutron-star equation of state (EOS) through the tidally induced shift in the waveform phase that depends on the tidal deformability parameter $\lambda$. Previous work has shown that $\lambda$, a function of the neutron-star EOS and mass, is marginally measurable by Advanced LIGO for a single event when including tidal information up to the merger frequency. In this work, we describe a method for stacking measurements of $\lambda$ from multiple inspiral events to measure the EOS. We use Markov chain Monte Carlo simulations to estimate the parameters of a 4-parameter piecewise polytrope EOS that matches theoretical EOS models to a few percent. We find that, for ``realistic'' event rates ($\sim 40$ binary neutron star inspiral events per year with signal-to-noise ratio $> 8$ in a single Advanced LIGO detector), combining a year of gravitational-wave data with the constraints from causality and recent high mass neutron-star measurements, the EOS above nuclear density can be measured to better than a factor of two in most cases. We also find that in the mass range $1M_\odot$--$2M_\odot$, the neutron-star radius can be measured to $\pm 1$~km and the tidal deformability can be measured to $\pm 1 \times 10^{36}$~g~cm$^2$~s$^2$. The overwhelming majority of this information comes from the loudest $\sim 5$ events. Current uncertainties in the post-Newtonian waveform model, however, lead to systematic errors in the EOS measurement that are as large as the statistical errors, and more accurate waveform models are needed to remove this error.
\end{abstract}

@@

[[Update the PACS numbers.]]

[[Should you describe briefly where each theoretical EOS came from or is it just not relevant? Maybe people would want to know why the ALF2 EOS is not fit very well but the others are.]]

@@

@@

[[How do I remove the little space before the underscore in \texttt{lalinference\_mcmc}?]]

[[State the value of $\rho_{\rm nuc}$]]

@@

\begin{figure}[!htb]
\begin{center}
\includegraphics[width=3.2in]{Radiuslambda.pdf}
\caption{Radius and tidal deformability of tabulated EOS models (solid) and the least-squares piecewise-polytrope fits to those tabulated models (dashed). The 20 vertical lines represent the most likely NS masses of the 10 known BNS systems~\cite{Lattimer2012}. Some of these masses, however, have significant uncertainties. The overlapping vertical bands represent the $1\sigma$ uncertainty in the masses of the pulsars J1614-2230 ($1.97\pm0.04M_\odot$)~\cite{DemorestPennucciRansom2010} and J0348+0432 ($2.01\pm0.04M_\odot$)~\cite{AntoniadisFreireWex2013}, both in neutron star-white dwarf binaries.}
\label{fig:rlofm}
\end{center}
\end{figure}

@@

\red{[[Make sure all the waveform parameters are explicitly defined in Section II, or else you'll have to define them here.]]}

Given the GW model and EOS fit described in Sections~\ref{sec:waveform} and~\ref{sec:eos}, we will describe here a method for estimating the EOS parameters from a set of GW observations. We use Bayes' theorem to write the posterior density function (PDF) for the universal EOS parameters $\vec E = (\log(p_1), \Gamma_1, \Gamma_2, \Gamma_3)$ that are common to all NSs and the waveform parameters $\vec\theta = (D_L, \alpha, \delta, \psi, \iota, t_c, \phi_c, m_1, m_2, \tilde\Lambda, \delta\tilde\Lambda)$ that are unique to each of the $n$ BNS systems as
\begin{equation}
\begin{split}
p(\vec E,\vec\theta_1,\dots,\vec\theta_n | D,\mathcal{H},\mathcal{I})
&= p(\vec E,\vec\theta_1,\dots,\vec\theta_n | \mathcal{H},\mathcal{I}) \\
& \times\frac{ p(D | \vec E,\vec\theta_1,\dots,\vec\theta_n,\mathcal{H},\mathcal{I}) }{ p(D | \mathcal{H},\mathcal{I}) }.
\end{split}
\end{equation}
The data $D = \vec d_1,\dots, \vec d_n$ is composed of the data streams $\vec d_i(t)$ from the GW detector network for each of the $n$ inspiral events. $\mathcal{H}$ represents the EOS and waveform models, and $\mathcal{I}$ represents the background information for the parameters. The quantity $p(\vec E,\vec\theta_1,\dots,\vec\theta_n | \mathcal{H},\mathcal{I})$ is the prior probability density for the EOS and waveform parameters, $p(D | \vec E,\vec\theta_1,\dots,\vec\theta_n,\mathcal{H},\mathcal{I})$ is the likelihood, and the normalization constant $p(D | \mathcal{H},\mathcal{I})$, which we will not need to calculate here, is the evidence. The marginalized PDF for the EOS parameters is then
\begin{equation}
\label{eq:margEOS}
p(\vec E | D,\mathcal{H},\mathcal{I}) = \int d\vec\theta_1 \dots d\vec\theta_n p(\vec E,\vec\theta_1,\dots,\vec\theta_n | D,\mathcal{H},\mathcal{I}).
\end{equation}
This $11n$-dimensional integral is not easily computed, so we will decompose it into blocks that can be individually computed. First, we decompose the waveform parameters into extrinsic parameters $\vec\theta_{\rm ex} = (D_L, \alpha, \delta, \psi, \iota, t_c, \phi_c, \delta\tilde\Lambda)$\footnote{$\delta\tilde\Lambda$ is an intrinsic parameter that in principle provides additional EOS information. However, it is unmeasurable with aLIGO~\cite{WadeCreightonOchsner2014}, so we treat it as a nuisance parameter and group it with the extrinsic parameters when we marginalize over them.} that depend on the observer and intrinsic parameters $\vec\theta_{\rm in} = (m_1, m_2, \tilde\Lambda)$ that are relevant for measuring the EOS such that $\vec\theta = (\vec\theta_{\rm ex}, \vec\theta_{\rm in})$.

@@

The marginalized PDF (Eq.~\eqref{eq:margEOS}) is now
\begin{equation}
\begin{split}
p(\vec E | D,\mathcal{H},\mathcal{I}) = &\frac{1}{p(D | \mathcal{H},\mathcal{I})} 
\int d\vec\theta_{\rm in, 1} \dots d\vec\theta_{\rm in, n} \left\{ p(\vec E | \mathcal{H},\mathcal{I}) \right. \\
& \times \prod_{i=1}^n \left[p(m_{1i}, m_{2i} | \vec E, \mathcal{H},\mathcal{I}) p(\tilde\Lambda_i | m_{1i}, m_{2i}, \vec E, \mathcal{H},\mathcal{I}) \right.\\
& \left.\left. \times L(\vec d_i; \vec\theta_{\rm in, i}, \mathcal{H}, \mathcal{I}) \right]\right\},
\end{split}
\end{equation}
where we have defined the quasilikelihood for the intrinsic parameters as
\begin{equation}
\label{eq:quasilike}
L(\vec d_i; \vec\theta_{\rm in, i}, \mathcal{H}, \mathcal{I}) = \int d\vec\theta_{\rm ex, i} p(\vec\theta_{\rm ex, i} | \mathcal{H}, \mathcal{I}) p(\vec d_i | \vec\theta_i,\mathcal{H},\mathcal{I}).
\end{equation}
Because $\tilde\Lambda_i$ is a deterministic function of $m_{1i}$, $m_{2i}$ and the EOS parameters,
\begin{equation}
p(\tilde\Lambda_i | m_{1i}, m_{2i}, \vec E, \mathcal{H},\mathcal{I}) = \delta(\tilde\Lambda_i - \tilde\Lambda(m_{1i}, m_{2i}, \vec E)).
\end{equation}
The marginalized PDF finally becomes
\begin{equation}
\label{eq:margEOS2}
\begin{split}
p(\vec E | D,\mathcal{H},\mathcal{I}) =& \frac{1}{p(D | \mathcal{H},\mathcal{I})} \int dm_{11} dm_{21} \dots dm_{1n} dm_{2n} \\
&\times p(\vec E | \mathcal{H},\mathcal{I}) \prod_{i=1}^n \left[ p(m_{1i}, m_{2i} | \vec E, \mathcal{H},\mathcal{I}) \right.\\
& \left. \times L(\vec d_i; \vec\theta_{\rm in, i}, \mathcal{H}, \mathcal{I})|_{\tilde\Lambda_i = \tilde\Lambda(m_{1i}, m_{2i}, \vec E)} \right].
\end{split}
\end{equation}
The problem has now been reduced to computing the quasilikelihood (Eq.~\eqref{eq:quasilike}) for each GW event, then computing Eq.~\eqref{eq:margEOS2}.

\subsection{Likelihood and signal to noise ratio}

The final ingredient we need to evaluate the marginalized PDF is an expression for the likelihood $p(\vec d_i | \vec\theta_i,\mathcal{H},\mathcal{I})$ for each GW event\footnote{In this subsection, which discusses the likelihood for individual GW events, we omit the event index $i$ for clarity.}. In this paper we assume that each detector in the network has stationary, Gaussian noise and that the noise between detectors is uncorrelated. This means that the power spectral density (PSD) $S_n(f)^a$ of the noise $n^a(t)$ in detector $a$ is \red{[[cite]]}
\begin{equation}
\langle \tilde n^a(f) \tilde n^{b*}(f') \rangle = \frac{1}{2} \delta^{ab} \delta (f-f') S_n(f)^a,
\end{equation}
where $\tilde n^a(f)$ is the Fourier transform of the noise of detector $a$, and $\langle\cdot\rangle$ represents an ensemble average. For a GW event with true parameters $\hat\theta$, resulting in the gravitational wave signal $h^a(t; \hat\theta)$, the data stream of detector $a$ will be
\begin{equation}
d^a(t) = n^a(t) + h^a(t; \hat\theta).
\end{equation}

For stationary, Gaussian noise, it is well known that the probability of obtaining the noise time series $n(t)$ is
\begin{equation}
p_n[n(t)] \propto e^{-(n, n)/2},
\end{equation}
where $(a, b)$ is the usual inner product between two time series $a(t)$ and $b(t)$ weighted by the PSD
\begin{equation}
(a, b)=4{\rm Re} \int_0^\infty \frac{\tilde a(f) \tilde b^*(f)}{S_n(f)}\,df.
\end{equation}
If we have a GW model $m(t;\vec\theta)$ that approximates the true signal $h(t;\vec\theta)$, then the likelihood of obtaining the data $d^a(t)$ given the GW model is
\begin{align}
p(d^a | \vec\theta, \mathcal{H}, \mathcal{I}) &= p_n[d^a(t)-m(t; \vec\theta)] \nonumber\\
& \propto e^{-(d^a - m(\vec\theta), d^a - m(\vec\theta))/2}.
\end{align}
If the model $m$ differs from the true signal $h$, a systematic error will be introduced in the recovered waveform parameters. For a network of independent detectors described by the set of time series $\vec d(t)$, the likelihood is then
\begin{equation}
p(\vec d | \vec\theta, \mathcal{H}, \mathcal{I}) = \prod_a p(d^a | \vec\theta, \mathcal{H}, \mathcal{I}).
\end{equation}

The observed signal to noise ratio (SNR) for the data $d^a$ from detector $a$ is Gaussian random variable given by
\begin{equation}
\rho^a = \frac{(d^a, m)}{\sqrt{(m, m)}},
\end{equation}
and has the expectation value (or characteristic SNR) $\langle \rho^a \rangle = \sqrt{(m, m)}$. The SNR for a network of detectors is then $\rho_{\rm net} = \sqrt{\sum_a(\rho^a)^2}$, which has expectation value $\langle \rho_{\rm net} \rangle = \sqrt{\sum_a\langle \rho^a \rangle^2}$.

@@

We evaluate Eq.~\eqref{eq:margEOS2} for the marginalized PDF in a two-step procedure, similar to that described by Steiner {\it et al.}~\cite{SteinerLattimerBrown2010}, where we first evaluate the quasilikelihood (Eq.~\eqref{eq:quasilike}) for each of the $n$ BNS systems, then evaluate Eq.~\eqref{eq:margEOS2}. In the first step where we evaluate the quasilikelihood $L(\vec d; \vec\theta_{\rm in}, \mathcal{H}, \mathcal{I})$, we use the MCMC sampler \texttt{lalinference\_mcmc} included in the LSC Algorithm Library~\cite{lal} as was implemented in Ref.~\cite{WadeCreightonOchsner2014} and is described in more detail in Ref.~\cite{VanDerSluysRaymondMandel2008}. We use a uniform prior distribution for $\log(D_L)$, $\alpha$, $\sin(\delta)$, $\psi$, $\sin(\iota)$, $t_c$, and $\phi_c$ for the extrinsic parameters. \red{[[The specific range is blah....]]} The priors for the intrinsic parameters of interest $(m_{1}, m_{2}, \tilde\Lambda)$ are not contained in the quasilikelihood (Eq.~\eqref{eq:quasilike}), so we are required to use flat priors. In most cases we used $1M_\odot \le m_2 \le m_1 \le 30M_\odot$ and $0 \le \tilde\Lambda \le 3000$. However, when we injected BNS systems with component masses of $1M_\odot$ in Section~\ref{sec:varymass}, we used $0.5M_\odot \le m_2 \le m_1 \le 30M_\odot$ and $0 \le \tilde\Lambda \le 5000$ so that the posterior was minimally affected by the prior. For $\delta\tilde\Lambda$, we used the range $-500 \le \delta\tilde\Lambda \le 500$\red{[[Is this true?]]}. For each binary event we perform 10 independent runs and test for convergence using the Gelman-Rubin diagnostic~\cite{GelmanRubin1992}, then join the samples. Marginalization over the extrinsic parameters is performed by generating a distribution from the samples of the three intrinsic parameters with a Gaussian kernel density estimator.

In the second step, we sample the $4+2n$ parameter integrand of Eq.~\eqref{eq:margEOS2} using the affine-invarient ensemble sampler \texttt{emcee}~\cite{ForemanMackeyHogg2013}. We use uniform priors for the EOS parameters with boundaries as described in Section~\ref{sec:eosparam} and additional boundaries from the $1.93M_\odot$ observation and causality constraints as described in Sections~\ref{sec:eosconstraints} and~\ref{sec:eosparam}, and uniform priors for the masses with $1M_\odot \le m_2 \le m_1 \le 3.0M_\odot$ or, when we examine $1M_\odot$ systems, $0.5M_\odot \le m_2 \le m_1 \le 3.0M_\odot$. We again perform 10 runs and test for convergence with the Gelman-Rubin diagnostic before joining the samples. The marginalized distribution $p(\vec E|D, \mathcal{H}, \mathcal{I})$ is then given by a histogram of the samples for $\vec E$.

The evaluation of the quasilikelihood (Eq.~\eqref{eq:quasilike}) for all $n$ events in a population is very computationally expensive, so in some cases we will use the Fisher matrix approximation instead. In the large SNR limit, the difference $\Delta\vec\theta = \vec\theta - \hat\theta$ between the estimated parameters $\vec\theta$ and the true parameters $\hat\theta$ of the binary system obeys a Gaussian distribution~\cite{FinnChernoff1993}. Specifically, for $N$ parameters, the likelihood is
\begin{equation}
\label{eq:gaussianlike}
p(d^a | \vec\theta,\mathcal{H},\mathcal{I}) = \frac{1}{\sqrt{(2\pi)^N \mathrm{det}(\Sigma_{ij})}} e^{-\frac{1}{2} \Sigma^{-1}_{ij} \Delta\theta_i \Delta\theta_j},
\end{equation}
where $\Sigma_{ij}$ is the covariance matrix, and it is given in terms of the Fisher matrix
\begin{equation}
\Gamma_{ij} = (\partial_i h(\hat\theta), \partial_j h(\hat\theta))
\end{equation}
by the relation $\Sigma_{ij}  = \Gamma^{-1}_{ij}$. In the large SNR limit, $\hat\theta$ will be approximately given by the maximum likelihood. When we use the Fisher matrix approximation we will use the mass variables $\ln(\mathcal{M})$ and $\ln(\eta)$ instead of $m_1$ and $m_2$ and flat priors for $\ln(\mathcal{M})$ and $\ln(\eta)$. The quasilikelihood (Eq.~\eqref{eq:quasilike}) marginalized over the extrinsic parameters is simply given by the submatrix of $\Sigma_{ij}$ containing the intrinsic parameters $(\ln(\mathcal{M}), \ln(\eta), \tilde\Lambda)$. However, when we estimate the EOS parameters in Eq.~\eqref{eq:margEOS2}, we always sample the posterior with \texttt{emcee}.

@@

In order to determine the characteristic ability of GW detectors to measure EOS parameters, we will want to average our results for Eq.~\eqref{eq:margEOS2} over many realizations of the detector noise. In Ref.~\cite{NissankeHolzHughes2010} it was shown that averaging over noise realizations is equivalent to setting the noise realization $n^a(t)$ used to generate the data $d^a(t)$ equal to zero, but still using the characteristic noise PSD in the expression for the likelihood. Explicitly, one can define the averaged likelihood as the geometric mean of an ensemble of $M$ identical detectors measuring the same GW event
\begin{equation}
p_{\rm ave}(d^a | \vec\theta, \mathcal{H}, \mathcal{I}) = \left[\prod_{k=1}^M p(d_k^a | \vec\theta, \mathcal{H}, \mathcal{I})\right]^{1/M}.
\end{equation}
Taking $M$ to be large, it was shown that
\begin{equation}
\label{eq:averagedlike}
p_{\rm ave}(d^a | \vec\theta, \mathcal{H}, \mathcal{I}) \propto e^{-\frac{1}{2}(m^a(\theta) - h^a(\hat\theta) | m^a(\theta) - h^a(\hat\theta))}.
\end{equation}
This averaged likelihood will allow us to examine characteristic measurement uncertainties in the EOS parameters independent of the particular noise realization. It will also allow us in Section~VI to separate the effects of systematic errors due to uncertainties in the waveform model from effects due to individual noise realizations.

@@

\begin{figure*}[!htb]
\begin{center}
\includegraphics[width=3.2in]{LALMCMCmpa1FitTaylorF2ZeroNoiseBNS20p.pdf}
\includegraphics[width=3.2in]{LALMCMCmpa1FitTaylorF2ZeroNoiseBNS20Radiuslambda.pdf}
\caption{Uncertainty in the recovered EOS, radius, and tidal deformability for the three-detector aLIGO--aVirgo network. Results are shown for the loudest 20 events with Network SNR from 64.1 down to 13.7. The red, green, and blue regions represent the 68\% ($1\sigma$), 95\% ($2\sigma$), and 99.7\% ($3\sigma$) confidence regions respectively. In the bottom left panel, $p_{\rm true}$ is the pressure of the true ``injected'' EOS, which in this case is the fit to the MPA1 EOS. In the right panels, the vertical line at $1.93M_\odot$ is the minimum mass, set by the prior, where some accepted EOS parameters do not produce a stable NS.}
\label{fig:mpa1fit}
\end{center}
\end{figure*}

@@

BL sincerely thanks Walter Del Pozzo and Tyson Littenberg for advice, many helpful discussions, and correcting misunderstandings in the implementation of Bayesian techniques, as well as Will Farr, Tjonnie Li, Chris Van Den Broeck, and Andrew Steiner. BL also had several informative discussions at the University of Washington's Institute for Nuclear Theory Program 14-2a. The authors would also like to thank Evan Ochsner and Jolien Creighton for advice and for generating the parameters of the BNS population used here. This work was supported by NSF Grants PHY-1305682, and PHY-\red{xxxxxxx [[Les, fill this in]]}. Simulations were run on the Nemo cluster at the University of Wisconsin--Milwaukee and the Orbital cluster at Princeton University.
